Repeatability of Pre-Flashover Fire Patterns on Gypsum Wallboard

### 2016-10-03 Projects/Madrzykowski_Thesis/Report/Madrzykowski_Thesis.tex
Added imaes to ch 6 and tables and text to Ch 7
cut out mccaffrey eqn to save

@@

\begin{sidewaystable}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Fuel (\# of tests)      &   Wall Type   &   Max. Height (m)	         & Max. Width (m)               &  Height at Max. Width (m) &  Area (m$^2$)  \\ \hline \hline
		Natural Gas (10)    &   Open 		& 0.74 	$\pm$ 0.12 (16~\%)   &	0.24 $\pm$ 0.06 (25~\%) 	& 0.41 $\pm$ 0.07 (17~\%)	& 0.15 $\pm$ 0.05 (33~\%) 	\\
		Natural Gas (10)    &   Closed 		& 0.74	$\pm$ 0.06 (8~\%)    &	0.24 $\pm$ 0.02 (7~\%) 		& 0.43 $\pm$ 0.03 (7~\%)  	& 0.14 $\pm$ 0.01 (6~\%) 	\\
		Natural Gas (10)    &   Insulated 	& 0.71	$\pm$ 0.10 (14~\%)   &	0.23 $\pm$ 0.04 (17~\%) 	& 0.42 $\pm$ 0.07 (17~\%) 	& 0.15 $\pm$ 0.02 (14~\%)  	\\
		Gasoline (10)       &   Open    	& 0.83	$\pm$ 0.15 (18~\%)   &	0.28 $\pm$ 0.09 (32~\%) 	& 0.44 $\pm$ 0.14 (32~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Closed  	& 0.80	$\pm$ 0.10 (13~\%)   &	0.29 $\pm$ 0.04 (14~\%) 	& 0.44 $\pm$ 0.06 (14~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Insulated	& 0.82 	$\pm$ 0.09 (11~\%)   &	0.25 $\pm$ 0.08 (32~\%) 	& 0.40 $\pm$ 0.13 (32~\%) 	& 0.13 $\pm$ 0.04 (30~\%)  	\\
		\hline
	\end{tabular}
	\caption[Comparison of fire pattern dimensions as a function of wall construction type]{Comparison of fire pattern dimensions as a function of wall construction type. Values presented with 95~\% confidence limits.}
	\label{tab:Fire_Pattern_Dimensions_Wall_Construction}
\end{sidewaystable}

@@

Figure~\ref{GBGas10Sequence} displays a sequence of photographs taken during a gasoline burner corner experiment.  The photographs taken at a rate of five per second show the significant movement of the regions of visible combustion.

@@

Table~\ref{tab:Fire_Pattern_Dimensions_Corners}

\begin{sidewaystable}
	\centering
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Fuel (\# of tests)      &   Wall Type   & Max. Height Left (m)	 	& Area Left (m$^2$)				&  Max. Height Right (m)        & Area (m$^2$)  				\\ \hline \hline
		Natural Gas (10)    &   Open 		& 0.74 	$\pm$ 0.15 (20~\%)  & 0.061 $\pm$ 0.028 (46~\%) 	&  0.78 $\pm$ 0.17 (22~\%) 	 	& 0.061 $\pm$ 0.028 (46~\%)  	\\
		Gasoline (10)       &   Open    	& 0.47	$\pm$ 0.26 (55~\%)  & 0.027 $\pm$ 0.019 (70~\%) 	&  0.45 $\pm$ 0.31 (69~\%)  	& 0.019 $\pm$ 0.011 (58~\%)\\
		PUF (3)       		&   Open 		& 0.32	$\pm$ 0.13 (41~\%)  & 0.056 $\pm$ 0.034 (61~\%) 	&  0.36 $\pm$ 0.091 (25~\%)  	& 0.071 $\pm$ 0.023 (32~\%)\\
	\hline
	\end{tabular}
	\caption[Comparison of corner fire pattern height and area for the three types of burners]{Comparison of corner fire pattern height and area for the three types of burners. Values presented with 95~\% confidence limits.}
	\label{tab:Fire_Pattern_Dimensions_Corners}
\end{sidewaystable}

@@

Figure~\ref{NG_Corner} shows a comparison of the ten fire patterns generated by the natural gas burner in the corner position with walls of open back construction.

@@

Figure~\ref{Gas_Corner} shows a comparison of eight of the ten fire patterns generated by the gasoline fueled burner in the corner position with walls of open back construction.

@@

Figure~\ref{PUF_Corner} shows a comparison of three fire patterns generated by the polyurethane foam fueled burner in the corner position with walls of open back construction. Only three replicates could be conducted due to the limits on the foam supply that was sourced from the same purchase.

@@

@@

Over the years, fire researchers have provided many methods for predicting flame characteristics such as: median flame height, plume temperatures and an estimation of the heat flux emitted. These methods range from simple, empirically based, algorithms to computational fluid dynamics models~\cite{Beyler:1986,Beyler:1999,FDS_Users_Guide,FDS_Tech_Guide}.

In this chapter the application of empirical algebraic flame characterizations will be compared against the results from the experiments documented in Chapter 2 and 4. Two different types of applications will be used.  In the first case relationships between the energy generated by a given fire source and the potential to transfer enough heat to generate thermal damage on a target, in this case gypsum board, will be examined.  In the second case, the height of the thermal damage on the gypsum board will be examined. Applications of algorithms such as these are presented in current fire dynamics and investigation textbooks~\cite{Icove:2013,Quintiere:1997,Asseal:2010,Beyler:1999}.

Values generated in the bench scale experiments will be utilized as input to predict or simulate real scale results using the algorithms.  Then the predicted results will be compared with the full-scale experimental results.

@@

The first algorithm assumes that the heat release rate is emitted from a point source in a uniform manner.  Therefore for any given radii a sphere of nearly uniform heat flux will be assumed. A schematic of this arrangement is shown in Figure~\ref{PointSourceHF} and the equation is presented below. Babrauskas applied this concept in his paper, ``Will the Second Item Ignite''~\cite{Babrauskas:1981}.  The idea being that if you knew the threshold heat flux that a material needed to be exposed to for the item to ignite.  Then given a fire with a specified HRR and radiant fraction, one could determine the range of distances from the fire where the material could receive enough energy to ignite.  Radiation is assumed to be the only means of heat transfer.

\begin{figure}
	\centering
	\includegraphics[width=\textwidth]{../Figures/PointSourceHF}
	\caption{Schematic of heat flux estimation using the point source assumption}
	\label{PointSourceHF},
\end{figure},

@@

\begin{tabbing}
where: \=  \\

\> $\dot{q}''$ is the heat flux per unit area kW/m$^2$\\
\> $\dot{Q}$ is the heat release rate (kW) \\ 
\> $\chi_r$ is the radiative fraction \\
\> $R$ = distance from the point source to the target (m) \\
\end{tabbing}

Source fire plume temp and shape factor from vertical cylinder
\begin{equation}\label{eq:source_fire}
\dot{q}'' = \epsilon \sigma F_{12} T^4  \\
\end{equation}

\begin{tabbing}
where: \=  \\
\> $\epsilon$ is the emissivity  \\
\> $\sigma$ is the Stefan-Boltzmann Constant (5.67 $\times 10^{−8}$~W/m$^2$~K$^4$)  \\
\> $F_{12}$ is the radiative shape factor \\
\end{tabbing}

@@

discuss the premise    ~\cite{McCaffrey:1979}
McCaffery

\begin{equation}\label{eq:z_c}
Z_c = 0.08 \dot{Q}^{\frac{2}{5}}
\end{equation}
where $Z_c$ is the continuous flame height~(m)
\begin{equation}\label{eq:z_i}
Z_i = 0.20 \dot{Q}^{\frac{2}{5}}
\end{equation}
where $Z_i$ is the intermittent flame height~(m)

Heskestad widely used  NRC CFITrainer  texts

\subsection{Correlation Descriptions}
Heskestad, developed an empirical method for calculating the mean visible flame height for fuels which do not have substantial in-depth combustion.  Appropriate fuels would ``include liquid pool fires and other horizontal surface fires'' add cites.  Given that the fuel configurations are square, a conversion of the fuel surface area to the effective diameter is required in order to use Heskestad's flame height equation directly.  Setting the area of the fuel surface, 0.093~m$^2$, equal to $\pi D^2/4$ and then solving for $D$ yields an effective diameter of 0.344~m.
Calculating the flame heights with the average peak heat release rates measured from the three fuels yields 1.0~m and 0.73~m for natural gas, gasoline, and polyurethane foam respectively. It should be noted that the coefficient, 0.235, is dependent on the ratio of the heat of combustion to the actual mass stoichiometric ratio of air to volatiles of the fuel.  For many liquid and gaseous fuels this coefficient is representative, Heskestad points out, however, that some common fuels, such as gasoline, deviate from the cited range. For gasoline a coefficient value of 0.2 should be used 32.          Taking the uncertainty of the measurements used to determine the heat release rate ($\pm$ 11~\%) into account generates a range of values.   The range of calculated values are compared with the mean flame height and range for flame heights  based on photographic analysis for each of the fuels in Table z.

@@

\section{Comparison of Results}

From Interflam  Screening algorithms utilizing the heat release rate results were employed to estimate the flame height and heat flux to a target surface.  These calculated values were compared to the measured results from “real fuels” and provided reasonable results.  Ranges of values were calculated in order to address the uncertainty in the heat release rate measurements.  Other uncertainties, such as lack of knowledge on the exact fuel composition need to be addressed, as this may well be the case in a real fire investigation.  Even accounting for uncertainties in heat release rate measurement, the flame height estimates tended to be better aligned with the peak flame height than the median flame height.  A specific coefficient for “gasoline” was used in Heskestad’s equation which resulted in the measured median flame height being within the range of the predictions.  The heat flux estimations were calculated with a range of radiant heat fractions, with the intent of the “actual” value being in that range.  With the exception of gasoline, the estimated heat flux values tended to under predict the measured results.  Given the limitations of the data, the algorithms provided reasonable estimates of the mean flame height and the heat flux.  These algorithms are tools which can be used to provide insight into a problem, but not an accurate single value answer.

@@

@@

The comparison for the fire pattern heights and flame heights from the natural gas fueled fires are very similar.  This is expected in part because the natural gas fires are “steady state”.  The “steady” nature of the fire aids in the repeatability or the reduction of variability in the experiments.  For example the energy transfer to the gypsum wall board is also fairly “steady” for the duration of the experiment.  Based on the average total heat released, approximately 23 MJ, more energy was transferred to the wall board when compared to the other two fuels.  This energy is needed to “dry” out the gypsum wallboard both of water absorbed from the environment and the chemically bound water, to allow the paper cover and paint to burn.

>>>>

The basic shape of these patterns on the closed wall are very similar to the patterns on the open wall.  Six of the ten patterns have a bump out or an unsymmetrical curve within 200~mm of the top of the burner.  The exact cause of these bumps in the patterns could not be determined, but they were generated later in the burn after the most of the fire pattern was created. Three of the patterns have a noticeable lean.

@@

\begin{sidewaystable}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Fuel (\# of tests)      &   Wall Type   &   Max. Height (m)	         & Max. Width (m)               &  Height at Max. Width (m) &  Area (m$^2$)  \\ \hline \hline
		Natural Gas (10)    &   Open 		& 0.74 	$\pm$ 0.12 (16~\%)   &	0.24 $\pm$ 0.06 (25~\%) 	& 0.41 $\pm$ 0.07 (17~\%)	& 0.15 $\pm$ 0.05 (33~\%) 	\\
		Natural Gas (10)    &   Closed 		& 0.74	$\pm$ 0.06 (8~\%)    &	0.24 $\pm$ 0.02 (7~\%) 		& 0.43 $\pm$ 0.03 (7~\%)  	& 0.14 $\pm$ 0.01 (6~\%) 	\\
		Natural Gas (10)    &   Insulated 	& 0.71	$\pm$ 0.10 (14~\%)   &	0.23 $\pm$ 0.04 (17~\%) 	& 0.42 $\pm$ 0.07 (17~\%) 	& 0.15 $\pm$ 0.02 (14~\%)  	\\
		Gasoline (10)       &   Open    	& 0.83	$\pm$ 0.15 (18~\%)   &	0.28 $\pm$ 0.09 (32~\%) 	& 0.44 $\pm$ 0.14 (32~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Closed  	& 0.80	$\pm$ 0.10 (13~\%)   &	0.29 $\pm$ 0.04 (14~\%) 	& 0.44 $\pm$ 0.06 (14~\%) 	& 0.17 $\pm$ 0.04 (24~\%)  	\\
		Gasoline (10)       &   Insulated	& 0.82 	$\pm$ 0.09 (11~\%)   &	0.25 $\pm$ 0.08 (32~\%) 	& 0.40 $\pm$ 0.13 (32~\%) 	& 0.13 $\pm$ 0.04 (30~\%)  	\\
		Polyurethane (10)	&   Open 		& 0.24  $\pm$ 0.12 (50~\%) 	 &  0.28 $\pm$ 0.08 (29~\%) 	& 0.40 $\pm$ 0.24 (60~\%)   & 0.05 $\pm$ 0.03 (57~\%)   \\
		\hline
	\end{tabular}
	\caption[Comparison of fire pattern dimensions as a function of wall construction type]{Comparison of fire pattern dimensions as a function of wall construction type. Values presented with 95~\% confidence limits.}
	\label{tab:Fire_Pattern_Dimensions_Wall_Construction}
\end{sidewaystable}

@@

Figure~\ref{GBGas10Sequence} displays a sequence of photographs taken during a gasoline burner corner experiment.  The photographs taken at a rate of five images per second show the significant movement of the regions of visible combustion, similar to the movement of the flames from the natural gas burner.

@@

The corner experiments demonstrated the expected extension of the combustion zone, especially in the cases of the natural gas and gasoline fires.  In the extreme case the flames from the gasoline burner briefly impinged on the ceiling, which is approximately 2.4~m above the top of the burner. The shape of the fire pattern to the wall was also different then the patterns generated on the flat wall.  Much samller areas of the wall had the painted paper charred, penetrated, and burned.  For each of the fuels there was significantly more area where it was difficult to separate the charred areas from the surface deposits and discoloration. The sectons below include post test images of the fire patterns.  The dimensions of the corner fire patterns are given in Table~\ref{tab:Fire_Pattern_Dimensions_Corners}.

\begin{sidewaystable}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
	Fuel (\# of tests)      &   Wall Type   & Max. Height Left (m)	 	& Area Left (m$^2$)				&  Max. Height Right (m)        & Area (m$^2$)  				\\ \hline \hline
		Natural Gas (10)    &   Open 		& 0.74 	$\pm$ 0.15 (20~\%)  & 0.061 $\pm$ 0.028 (46~\%) 	&  0.78 $\pm$ 0.17 (22~\%) 	 	& 0.061 $\pm$ 0.028 (46~\%)  	\\
		Gasoline (10)       &   Open    	& 0.47	$\pm$ 0.26 (55~\%)  & 0.027 $\pm$ 0.019 (70~\%) 	&  0.45 $\pm$ 0.31 (69~\%)  	& 0.019 $\pm$ 0.011 (58~\%)\\
		PUF (3)       		&   Open 		& 0.32	$\pm$ 0.13 (41~\%)  & 0.056 $\pm$ 0.034 (61~\%) 	&  0.36 $\pm$ 0.091 (25~\%)  	& 0.071 $\pm$ 0.023 (32~\%)\\
	\hline
	\end{tabular}
	\caption[Comparison of corner fire pattern height and area for the three types of burners]{Comparison of corner fire pattern height and area for the three types of burners. Values presented with 95~\% confidence limits.}
	\label{tab:Fire_Pattern_Dimensions_Corners}
\end{sidewaystable}

@@

Figure~\ref{NG_Corner} shows a comparison of the ten fire patterns generated by the natural gas burner in the corner position with walls of open back construction. In the corner burns there is an area in the corner where the paper is not burned off.  From the flame sequence shown in Figure~\ref{GBNG34Sequence} it can be seen that the flame in the persitent flame region is close to the walls of the corner, but moving up into the intermittent region, the flame burns away from the walls, hence less energy is transferred to the walls and perhaps a bit more soot condenses out on the cooler walls.

@@

Figure~\ref{Gas_Corner} shows a comparison of eight of the ten fire patterns generated by the gasoline fueled burner in the corner position with walls of open back construction. Due to a camera malfunction there were no comparable images for the other two experiments.  The patterns are darker than the patterns generated on the flat walls.  Given that the air to the combustion zone was constrained it follows that the combustion may have been less efficient than the flat wall experiments whihch resulted in the sootier patterns.  The combustion zone for the gasoline fires positioned in the corner stretched from the top of the burner up to the ceiling of the compartment.  Figure~\ref{GBGas10Sequence} shows the flame extension and similar to the natural gas burner fires shows how the flame in the intermittant zone appeared to move away from the corner.

@@

Figure~\ref{PUF_Corner} shows a comparison of three fire patterns generated by the polyurethane foam fueled burner in the corner position with walls of open back construction. Only three replicates were conducted due to the limits on the foam supply that was sourced from the same purchase.  The flame length from the polyurethane foam fueled burner was also longer compared to the flat wall experiments.  However the change was not as big compared to the higher heat release rate natural gas and gasoline fueled burners.  The resulting patterns, similar to the other corner burner patterns, were darker in color with less definition in the lines of demarcation.

@@

After the experiments the gypsum wallboard panels were removed from the compartment for documentation and measurement of the fire patterns.  A comparsion on a fire pattern from each type of burner is shown in Figure~\ref{PT_Corners}.  Another image of the gasoline generated corner fire pattern, Figure~\ref{C_PTGASClose}, is included to show the reader how viewing a fire pattern at different angles enabled a better sense of the area of the fire damage.

\begin{figure}[p]
	\includegraphics[width=1.7in]{../Figures/C_PTGBNG_42}
	\includegraphics[width=1.7in]{../Figures/C_PTGBGAS_51_10c}
	\includegraphics[width=1.7in]{../Figures/C_PTGBPUF_11} \\

	\caption[Photographs of the fire patterns from three different burner experiments positioned in a corner]{Photographs of the fire patterns from three different burner experiments positioned in a corner with walls of open construction.  The patterns were generated by the following fuels, from are from left to right, natural gas, gasoline, polyurethane foam.}
	\label{PT_Corners}
\end{figure}

\begin{figure}[h]
	\includegraphics[width=5in]{../Figures/C_PTGASClose} \\

	\caption[Photographs of a gasoline generated corner fire pattern taken at an angle]{Photographs of a gasoline generated corner fire pattern taken at an angle in order to improve the visibility of the lines of demarcation.}
	\label{C_PTGASClose}
\end{figure}

In general, the experiments, where the the burners were positioned against the center of a flat wall, provided very repeatable results with the natural gas and gasoline fueled burners. Under these pre-flashover, short burn duration fire conditions it was clear that the wall construction did not have any impact on the resulting fire pattern.  Also comparing these patterns to those generated with the instrumented gypsum wallboard, it did not appear that the instrumentation had any impact that changed the fire pattern.  With an expanded uncertainty of 2~$\sigma$, the height of the fire patterns were within $\pm$ 18~\% for both the natural gas and the gasoline fuel burner fires.  The uncertainties regarding the width and area of the patterns were higher up to $\pm$ 33~\%.  While the natural gas and the gasoline are diffent fuels there was no significant difference in any of the measured pattern dimensions between the two as show in Table~\ref{tab:Fire_Pattern_Dimensions_Wall_Construction}.  The larger uncertainty in the areas of the pattern may be due to the slight movement of the flames back and forth across the wall.

The comparison for the fire pattern heights and flame heights from the natural gas fueled fires are very similar.  This is expected in part because the natural gas fires are “steady state”.  The “steady” nature of the fire aids in the repeatability or the reduction of variability in the experiments.

The gasoline fueled fires also had similar fire pattern heights and flame heights; however, the similarity was not as strong as the natural gas fires.  The transient nature of the fire was believed to have had an impact since there was less energy transferred to the gypsum wall board and the exposure time for portions of the wall board more distant from the fuel surface was less.

The polyurethane foam exhibited a higher variability in the fire pattern heights, $\pm$ 50~\%, and areas, $\pm$ 57~\%, due in part to variabilities in fire growth rates, a lower peak heat release rate, and a lower total heat release, than the other two fuels. Since the fire patterns generated by the polyurethane fueled burners were shorter and smaller, than the patterns generated by the 80 kW heat release rate fuels, the uncertainty in terms of percentage difference is magnified.  The positive message is that the repeatability of the patterns qualitatively agree with the repeatability, or lack there of, of the heat release rate of the polyurethane foam burner which was $\pm$ 47~\% during the burner characterization experiments.

@@

Over the years, fire researchers have provided many methods for predicting flame characteristics such as: median flame height, plume temperatures and an estimation of the heat flux emitted. These methods range from simple, empirically based, algorithms to computational fluid dynamics models~\cite{Beyler:1986,Beyler:1999,Heskestad:SFPE,FDS_Users_Guide,FDS_Tech_Guide}.

In this chapter the application of empirical algebraic flame characterizations will be compared against the results from the experiments documented in previous chapters. Two different types of applications will be used.  In the first case relationships between the energy generated by a given fire source and the potential to transfer enough heat to generate thermal damage on a target, in this case gypsum board, will be examined.  In the second case, the height of the thermal damage on the gypsum board will be examined. Applications of algorithms such as these are presented in current fire dynamics and investigation textbooks as well as other application oriented documents~\cite{Icove:2013,Quintiere:1997,Asseal:2010,Beyler:1999,NUREG_1805}.  Values generated in the fire characterizations, instrumented bench scale experiments, and values from the literature will be utilized as input to predict or simulate real scale results using the algorithms.  Then the predicted results will be compared with the full-scale experimental results.  keep in mind that with these algorithms time is not a factor.  Methods of this type are commonly referred to as ``back of the envelope'' estimates.  they are intended to provide insight as opposed to a hard and fast numerical answer.  As the experiments have shown so far there are many uncertainities which impact the results.

@@

The first algorithm assumes that the heat release rate is emitted from a point source in a uniform manner.  Therefore for any given radii a sphere of nearly uniform heat flux will be assumed. A schematic of this arrangement is shown in Figure~\ref{PointSourceHF} and the equation is presented below. Babrauskas applied this concept in his paper, ``Will the Second Item Ignite''~\cite{Babrauskas:1981}.  The idea being that if you knew the threshold heat flux that a material needed to be exposed to for the item to ignite.  Then given a fire with a specified HRR and radiant fraction, one could determine the range of distances from the fire where the material could receive enough energy to ignite.  Radiation is assumed to be the only means of heat transfer.  However this method is noted to work the best when the distance of the target is at least two to four times the distance of the flame height way from the fire.

\begin{figure}
	\centering
	\includegraphics[width=4.5in]{../Figures/PointSourceHF}
	\caption{Schematic of heat flux estimation using the point source assumption}
	\label{PointSourceHF},
\end{figure},

@@

\begin{tabbing}
where: \=  \\
{}
\> $\dot{q}''$ is the heat flux per unit area kW/m$^2$\\
\> $\dot{Q}$ is the heat release rate (kW) \\ 
\> $\chi_r$ is the radiative fraction \\
\> $R$ = distance from the point source to the target (m) \\
\end{tabbing}

The question is can we use this simple model to estimate the burner positions where the heat flux would be high enough to generate thermal damage on the painted gypsum board wall?  From the source fire characterizations, the average peak heat release rates from the three burners can be input into the equation along with the distance, R, and the $\chi_r$, or the radiant fraction.  R is measured from the vertical centerline of the burners so in addition to the distance between the edge of the burner and the wall, half of the burner width, 0.15~m must be added.  A typical default value used for the radiant fraction is 0.35.  Table~\ref{tab:Point_Source_HFNGG} lists the input value of the average peak heat release rate of 80~kW for both the natural gas and gasoline burner fires along with a $\chi_r$ of 0.35, and the five burner positions that were used.

The predictions provide good agreement with the natural gas results at distances greater than 1 burner width away, then the predictions trend higher.  In the instrumented gypsum wallboard experiments the natural gas burner at $R=0.31~m$ generated a discoloration on the surface of the painted gypsum wallboard.  At $R=0.15~m$, burner touching the wall, enough energy was transferred to the wall fast enough that caused charring, penetration, and consumption of the painted paper surface.  Based on the data from the cone calorimeter experiments conducted as part of this study, charring, penetration, and consumption would occur somewhere between 25 and 30~kW/m$^2$.  Dillon provided a value of 26kW/m$^2$ as the minimum heat flux where ignition of the gypsum wallboard would occur~\cite{Dillon:1998}. So at $R=0.31~m$ conditions are on the edge both with the prediction and the measurement of ignition but not high enough to cause the paper to char or worse, there was only enough energy to cause discoloration.  Once the natural gas burner was moved next to the wall, the prediction and the measurement, even though they are not quantatively in agreement, both provide a heat flux in excess of 26~kW/m$^2$.  Indicating that the a fire pattern has been burned into the wall and that was infact true.

The predictions provided in Table~\ref{tab:Point_Source_HFNGG} are significantly lower than the measured peak heat flux values from the gasoline fueled burner experiments at every burner position.  That said, the predictions indicate that at $R=0.31~m$  the heat flux is just before the ignition threshold provided in the literature and at $R=0.23~m$ or less the heat flux is high enough to indicate ignition of the gypsum wallboard.  The single gasoline experiment at $R=0.31~m$ showed a peak heat flux of 57~kW/m$^2$.  The exposure was brief at this peak due to the flame movement toward the wall.  This makes the experimental values at the three closest location to the wall difficult to compare as the point source model, and most models for that matter, can not account for the flame being drawn toward the wall.  An important observation is that the fire damage on the wall was limited to an uneven horseshoe shaped pattern within 100~mm above the top of the burner edge.  This ignition was captured in Ch 4, Figure~\ref{IWGBGAS6_seq}. At the two closer positions the char, penetration, and consumption was evident and at these locations both the predictions and the measured heat fluxes are well above the threshold ignition heat flux.

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|c|}
	\hline
		HRR (kW) &   $\chi_r$   & R (m)	 & Heat Flux (kW/m$^2$)     &  Nat. Gas (kW/m$^2$)      & Gasoline (kW/m$^2$) 	\\ \hline \hline
		80    &   .35 		& 0.76 	     & 3.8 $\pm$ 0.30 (8~\%) 	&  4 $\pm$ 0.32 (8~\%) 	 	   & 6 $\pm$ 0.5 (8~\%)     \\
		80    &   .35    	& 0.46	     & 10.6 $\pm$ 0.80 (8~\%) 	&  11 $\pm$ 0.88 (8~\%)  	   & 18 $\pm$ 1.4 (8~\%)    \\
		80    &   .35 		& 0.31	     & 24.0 $\pm$ 1.9 (8~\%) 	&  23 $\pm$ 1.8 (8~\%)  	   & 57 $\pm$ 4.6 (8~\%)     \\
        80    &   .35 		& 0.23	     & 43.0 $\pm$ 3.4 (8~\%) 	&  NA 					  	   & 60 $\pm$ 4.8 (8~\%)     \\
        80    &   .35 		& 0.15	     & 95.0 $\pm$ 7.6 (8~\%) 	&  55 $\pm$ 4.4 (8~\%)  	   & 61 $\pm$ 4.9 (8~\%)     \\
	\hline
	\end{tabular}
	\caption[Comparison of calculated point source values (HRR = 80~kW) with heat flux data from the instrumented gypsum boardwall experiments]{Comparison of calculated point source values with heat flux data from the natural gas and the gasoline instrumented gypsum boardwall experiments. Values presented with 95~\% confidence limits based on the estimated uncertainty of heat flux values from the literature~\cite{Pitts:2006}.}
	\label{tab:Point_Source_HFNGG}
\end{table}

Table~\ref{tab:Point_Source_HFPUF} displays a similar development based on the average peak heat release rate,45~kW, for the polyurethane foan burner.
Similar to the results from the gasoline comparison, the predictive heat flux values at all of the burner positions are low relative to the experimental data.  But again we will examine the results along a more qualitative basis. The experimental data provides an average peak value of 25~kW/m$^2$ for $R=0.31~m$. For the polyurethane foam burns the peak values were only averaged over a 15 second range due to the brief nature of the peak heat release rate. At $R=0.31~m$ there was no charring of the painted paper on the gypsum wallboard, only soot deposition.  No thermal damage.  At at $R=0.23~m$ or less, the heat flux is high enough to indicate ignition in the experiments and that was the result.  Only the prediction at $R=0.15~m$ would definitely indicate that the gypsum wallboard would ignite based on the threhold heat flux.

\begin{table}
	\centering
	\footnotesize
	\begin{tabular}{|c|c|c|c|c|}
	\hline
		HRR (kW) &   $\chi_r$   & R (m)	     & Heat Flux (kW/m$^2$)     &  Polyurethane (kW/m$^2$)    	\\ \hline \hline
		45       &   .35 		& 0.76 	     & 2.1 $\pm$ 0.17 (8~\%) 	&  3.5 $\pm$ 0.28 (8~\%) 	   	 \\
		45       &   .35    	& 0.46	     & 6.0 $\pm$ 0.48 (8~\%) 	&  12.3 $\pm$ 0.98 (8~\%)  	     \\
		45       &   .35 		& 0.31	     & 13.5 $\pm$ 1.1 (8~\%) 	&  25 $\pm$ 2.0 (8~\%)   	      \\
        45       &   .35 		& 0.23	     & 24.1 $\pm$ 1.9 (8~\%) 	&  36 $\pm$ 2.9 (8~\%)		      \\
        45       &   .35 		& 0.15	     & 53.5 $\pm$ 4.3 (8~\%) 	&  59 $\pm$ 4.7 (8~\%)   	       \\
	\hline
	\end{tabular}
	\caption[Comparison of calculated point source values (HRR = 45~kW) with heat flux data from the instrumented gypsum boardwall experiments]{Comparison of calculated point source values with heat flux data from the polyurethane foam burner instrumented gypsum boardwall experiments. Values presented with 95~\% confidence limits based on the estimated uncertainty of heat flux values from the literature~\cite{Pitts:2006}.}
	\label{tab:Point_Source_HFPUF}
\end{table}

This was an interesting exercise but was not definitive in any way.  If we were to attempt to improve the predictions by using radiative fractions that are known to be representative of the fuels, the predictions in some cases might get worse.  For example, recommended values in the literature for the $chi_r$ of natural gas range from 0.15 to 0.25 for similar sized burners~\cite{Hostikka:3,McCaffrey:1982}.  The default radiative fraction of 0.35 is inline with the values found in the literature for gasoline~\cite{Hamins:1991,McGrattan:2000}.  Literature values for polyurethane foam were elusive.

One more way to assess the point source model was to examine the heat flux data measured at 2~m away from the vertical centerline on the burners during some of the early source fire characterization experiments. If we rearrange the point source equation, to solve for $\chi_r$ by using the measured average peak heat release rates and average peak heat fluxes at a distance of 2~m away the resulting value for $\chi_r$ are 0.26, 0.46 and 0.33 for the natural gas, gasoline and polyurethane burners respectively.  This distance should be more appropriate for the use the point source model but the resultant radiative fractions for the natural gas and gasoline seem high compared to the values found in the literature.  It appears that more research is needed on radiative fractions of fire sizes and realistic fuels that might be found in a residential setting

@@

The desire to use flame height correlations is based on the thought that the flame height will correlate to the height of the associated fire pattern.
With the heat release rate data that we have from these experiments, comparisons will be made with the one of the more widely used empirical algorithms, known as the Heskestad flame height correlation.

\subsection{Correlation Description}

Heskestad, developed an empirical method for calculating the mean visible flame height for fuels which do not have substantial in-depth combustion.  Appropriate fuels would include liquid pool fires and other horizontal surface fires~\cite{Heskestad:SFPE,Beyler:1986}. The inputs required to use the model are the heat release rate and the diameter of the fire or the burner.

@@

Given that the burner used in this study are square, a conversion of the fuel surface area to the effective diameter is required in order to use Heskestad's flame height equation directly.  Setting the area of the fuel surface, 0.093~m$^2$, equal to $\pi D^2/4$ and then solving for $D$ yields an effective diameter of 0.344~m.

Once again the average peak heat release rates from Ch 2 will be used.  The natural gas and the gasoline fueled burners both had an average peak heat release rate of 80 kW.  The polyurethane fueled burner had an average peak heat release rate of 45 kW.  Calculating the median flame heights with the average peak heat release rates measured from the three fuels yielded 1.0~m, 1.0~m, and 0.73~m for natural gas, gasoline, and polyurethane foam respectively. It should be noted that the coefficient, 0.235, is dependent on the ratio of the heat of combustion to the actual mass stoichiometric ratio of air to volatiles of the fuel.  For many liquid and gaseous fuels this coefficient is representative, Heskestad points out, however, that some common fuels, such as gasoline, deviate from the cited range. For gasoline a coefficient value of 0.2 should be used~\cite{Heskestad:SFPE}. Changing the coefficient yields a median flame height of 0.8~m for gasoline.

Based on the video based flame height analysis conducted as part of this study the median flame heights were 0.705~m, 0.695~m, and 0.470~m for natural gas, gasoline, and polyurethane, respectively.    Taking the uncertainty of the measurements used to determine the heat release rate ($\pm$ 11~\%) into account generates a range of values.   The range of calculated values are compared with the mean flame height and range for flame heights  based on photographic analysis for each of the fuels in Table z.

\begin{table}
\centering
\footnotesize
\begin{tabular}{|c|c|c|c|}
\hline
Method          &   \multicolumn{3}{|c|}{Flame Heights (m)} \\ \cline{2-4}
                &       Natural Gas	            &   Gasoline	            & Polyurethane Foam \\ \hline \hline
Photographic    &       0.70 $\pm$ 0.23 (33~\%)   &	0.84 $\pm$ 0.28 (33~\%) 	& 0.46 $\pm$ 0.23 (50~\%)   \\
Video Based     &   	0.70 $\pm$ 0.11 (15~\%)   &	0.70 $\pm$ 0.1 (14~\%) 	& 0.47 $\pm$ 0.16 (34~\%)   \\
Fire Pattern    &   	0.74 $\pm$ 0.12 (16~\%)   &	0.83 $\pm$ 0.15 (18~\%) 	& 0.24 $\pm$ 0.12 (50~\%)   \\
Heskestad (range)  &   	0.94 to 1.06		    &	0.65 to 0.93        	& 0.48 to 0.90   \\
\hline
\end{tabular}
 \caption[Comparison of the median flame height measurement results with the fire pattern heights and the Heskestad predictions]{Comparison of the median flame height measurement results with the fire pattern heights and the Heskestad predictions. The measured results are shown with 95~\% confidence limits.  The Heskestad median flame height calculations are presented as a range based on the uncertainies of the peak heat release rate of each fuel type.} 
 \label{tab:Comparing_Flame_Heights}
\end{table}

@@

Similar to the previous chapter the goal is to utilize an analytical tool to predict or simulate the results from some of the key experiments conducted in this study. Unlike the previous chapter the computational fluid dynamics (CFD) models are not as widely used by fire investigators as the algebraic tools are.  The CFD models require more knowledge and more input data to utilize correctly.  Currently modelers need to develop there own data or rely on ill defined databases with the hope that the material tested is similar to the materials of interest.

@@
